---
title: "On the distribution of yeasts of the genus _Mrakia_ in foods"
author: "E. Parente, DAFE, Università degli Studi della Basilicata"
date: "2026-06-06"
output:
  html_document:
    toc: yes
    toc_float: yes
    df_print: paged
bibliography: 
# uncomment the following line and change the name of the project specific bibliography
#  - My_refs.bib
# do not change this line
  - R_citations.bib
  
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl  
nocite: |
  @*
---
<!--
# READ ME  
This document is built over the where_is_my_bug code available in [FoodMicrobionet repository](https://github.com/ep142/FoodMicrobionet/tree/master/WIMB) and originally used for [this paper](https://www.sciencedirect.com/science/article/abs/pii/S0168160523000405). 
Code was revised using Claude AI (Sonnet v 4.6)
The script uses several functions which are sourced from the "source" folder. This is motivated by the fact that I re-use these functions in other script and this way is simpler to keep them up to date.
The p (parameters) list basically controls what the document can do. Read and change as appropriate before rendering. A few examples:

* if you want to search for a single genus put the genus name in p$taxa_to_search, set p$taxon_level to genus and p$go_down_to to species  

* if you want to search for two or more possibly unrelated taxa put them as a vector (example c("Lactobacillus", "Levilactobacillus") and set p$taxon_level and p$go_down_to to genus: avoid using more than 6-8 genera

* if you want to search for all genera in a family put the family name in p$taxa_to_search, set p$taxon_level to family and p$go_down_to to genus. Note that you might have too many taxa to plot

The other options should be straightforward after you play with them a little bit. In this version I am not going below L1 but, technically, when working with a family, you could select a single genus and do the plots for all L4 levels in which it appears. Of course this can be done in a loop.
**Note**: this version can work with either bacteria or fungi but not both at the same time: you can control it with p$domain: this must be coherent with p$taxa_to_search: I have not included error trapping for this, may be I will do it in the future.  
-->
<!-- 
## Notes for self (development)

This is a development version. I need to transform it in a .qmd document and make it more general and may be interactive.  

* need to replace phylotools with another package to read fasta files, need to decide the framework I want to use. 

* here I am using "species" for taxa names; to make it more general it should simply be taxon after the pooled_edges_sample_study df is created; which taxon level it is is clear from p$taxon_level
-->

```{r setup, include=FALSE}

# Install/load packages ---------------------------------------------------
knitr::opts_chunk$set(echo = FALSE)


.cran_packages <- c("tidyverse", "data.table", "bipartite", 
                    "knitr", "kableExtra")


# Fast check for missing packages
# find.package() is faster than installed.packages() for large libraries
.is_installed <- function(pkg) {
  isTRUE(length(find.package(pkg, quiet = TRUE)) > 0)
}

# Install/load pak
if(!require("pak", quietly = T)) {
  install.packages("pak")
  require(pak)
}

# Handle CRAN
.missing_cran <- .cran_packages[!sapply(.cran_packages, .is_installed)]
if(length(.missing_cran) > 0) {
  pak::pkg_install(.missing_cran)
}
.loaded <- sapply(.all_packages, require, character.only = TRUE)
if (any(!.loaded)) message("Failed to load: ", paste(names(.loaded)[!.loaded], collapse = ", "))


# Install/load functions
# this script uses several functions which must be available in the source folder
# if they are not, rendering will stop
# Directory and file checks
if (!dir.exists("source")) stop("The 'source' directory does not exist")

needed_functions <- c("get_FMBN.R", "make_distplot.R", "make_dotplot.R", 
                      "WIMB_utils.R", "update_r_citations.R", 
                      "get_function_dependencies.R")
source_functions <- list.files("source")

missing_functions <- setdiff(needed_functions, source_functions)

if (length(missing_functions) > 0) {
  stop("Missing functions from the source folder: ", paste(missing_functions, collapse = ", "))
}
# silently source the files using walk()
purrr::walk(needed_functions, \(x) source(file.path("source", x)))

# grab user defined functions from the environment
# This automatically excludes get_function_dependencies and update_r_citations
# by looking for them specifically or running right after your main scripts are loaded.
all_user_functions <- lsf.str()
my_functions <- setdiff(all_user_functions, c("get_function_dependencies", "update_r_citations"))
packages_in_functions <- get_function_dependencies(my_functions)

all_packages  <- unique(c(packages_in_functions, .cran_packages))
# generate your completely hands-off .bib bibliography file
update_r_citations(packages = all_packages, quiet = TRUE)

# Load packages into session, and print package version
invisible(sapply(all_packages, 
                 \(x) require(x, character.only = TRUE, quietly = TRUE)))

```

```{r options, message=F, warning= F}
# technically, it would be safer to recalculate each time because I might run
# the script from different starting points
# for reproducibility reasons
set.seed(100)
# mainly for reproducibility reasons (this will be saved with the work space)
r_version <- R.Version() # in the future may be check that the version running is compatible
# with the script (it only matters if somebody other than myself is using the script)
session_info <- sessionInfo()

# parameters, as a list

p<- list(
  # options for opening FoodMicrobionet
  # the first two are the current location of the latest version of FoodMicrobionet
  FMBN_repo = "ep142/FoodMicrobionet/",
  FMBN_path = "29db4568f9dbcbc1ebc39a44b7b538fd708f4950/the_real_thing/FMBN_5_1/FMBN_plus.rds",
  use_local = TRUE,
  force_download = FALSE,
  do_cache = TRUE,
  # primary output options
  out_folder = "WIMB_v2", # the output folder
  verbose = T, # F suppresses part of the output
  save_data = T, # saves tables and other data
  save_graph = T, # saves graphs
  # graph saving
  graph_type = ".jpeg", # graph device
  graph_res = 150, #  resolution for saved graphs
  graph_width = 7, # graph width, inches
  graph_height = 7, # graph height inches
  out_fn_pref = "Mrakia_s_", # prefix for saved files
  # options for taxonomic filtering
  taxon_level = "genus", # choice is between "order", "family", "genus"
  go_down_to = "species", # "genus" or "species", but many taxa are not identified at species level
  keep_NA_lowest = T, # keep taxa which have NA at go_down_to level
  domain = "Fungi", # Bacteria or Fungi, this is auto-detected by the script
  taxa_to_search = c("Mrakia"), # the taxa to search, can be a vector
  # options for removing Chloroplast and/or Mitochondria and and/or poorly identified taxa
  remove_chloroplast = F,
  remove_mitochondria = F,
  remove_no_phylum = T,
  # sample filters
  s_type = "Sample", # Sample or Environment or both
  food_grouping_variable = "L1", # L1 or L4; avoid L6 too narrow
  min_read_length = 50, # minimum length of sequences (it is an average by study) 
                        # will be overriden if domain == "Fungi"
  # a very low value means no filtering
  run_further_filters = F, # set T to apply additional filters by sequencing target/region
  gene_targets = c("ITS1", "ITS2", "ITS1-ITS2"), # restrict the gene targets
  # most frequent values for bacteria are V1-V3, V3-V4 and V4; for fungi ITS1 and ITS2
  log_base = 10, # log base for relative abundance transformation should be 2 or 10
    # plot layout and style
  plot_type = "boxplot",   # "boxplot" or "violin"
  flip_box_plot = T, # flip the axes in box plots
  violin_min_n = 10,       # fall back to boxplot if any group has fewer obs
  plot_facet_by    = "taxon",     # "food": taxon on x, food on facets; "taxon": food on x, taxon on facets
  plot_fill_var    = "rel_prev", # boxplot fill or point colour: "rel_prev", "spoilage", or "none"
  plot_show_jitter = TRUE,       # show individual sample points (grey, low alpha)
  plot_notch       = TRUE,       # notch the boxplot?
  # dot plot options
  dotplot_max_taxa      = 10,   # cap on number of taxa shown
  dotplot_max_foods     = 12,   # cap on number of food categories shown
  dotplot_size_var      = "median_logab",  # "median_logab" or "median_ab"
  # options for specialisation and ordination
  min_n_reads           = 5000, # minimum sequencing depth for specialisation/ordination
                                # samples below this threshold are excluded from d' and PCoA
                                # but NOT from prevalence/abundance summaries shown earlier;
                                # 10000 is a reasonable default for most amplicon studies; lower
                                # values will help including all F454 studies
  min_ab_specialisation = 0.001,# minimum relative abundance (proportion) for a taxon x food
                                # combination to be included in d' and PCoA (0.001 = 0.1%);
                                # guards against sequencing noise being treated as presence
  min_prev_specialisation = 3,  # minimum number of samples in which a taxon must appear
                                # in a food category to be included in d' and PCoA;
                                # guards against singletons driving the calculation
  bipartite_weight_var = "rel_prev", # "rel_prev" or "median_ab" but the latter is problematic for several reasons
  depth_weighted_prev = TRUE,  # if TRUE, weight each presence by min(1, n_reads2/min_n_reads)
                                # when computing prevalence for d';
                                # down-weights detections from shallow samples
  # miscellaneous
  abbr = 15                     # used to control length of abbreviated taxa and food 
                                # in graphs
)

# a manual color scale for T/F
TF_colors <- c("#F8766D", "#619CFF")
names(TF_colors) <- c("FALSE", "TRUE")
```

# What is this report

This is a quick and dirty report on the occurrence and (relative) abundance of `r str_flatten_comma(p$taxa_to_search)` in foods using [FoodMicrobionet version 5.1.2](https://github.com/ep142/FoodMicrobionet). Over time I will transform this in a Quarto document (a dashboard with a simplified workflow is already available).  
** Please note **: be very careful in not over-interpreting your results. Metataxonomic data are compositional and, when using FoodMicrobionet, you are grouping together several studies in which the absolute total abundance of a given group (bacteria or fungi) may be very different. Additionally, sequencing depth and target genes length are both important issues. Sequencing depth for samples in FoodMicrobionet may range over 3 orders of magnitude (from a few hundreds to >100,000). While in my experience 10,000 sequences are generally more than enough to uncover all the (relevant) diversity in foods, the probability of detecting rare species decreases with decreasing sequencing depths. These issues must be acknowledged and I have provided some options and guards in the parameter list.  


Here, I will be loading FMBN directly from GitHub (the last stable version): a cached copy will be stored and refreshed as needed.  

## R packages used in this analysis.  

`r base::version$version.string` [@R-base] was used for this report. I have used packages of the `tidyverse` [@R-tidyverse] and `data.table` [@R-data.table] for data wrangling and plotting. Package `bipartite` [@R-bipartite] was used to calculate specialization indices. Table generation in reports was handled with `knitr` and `kableExtra` [@R-knitr; @R-kableExtra]

## Loading data.  

```{r load-data}

local_input_path <-  NULL
# set this to NULL if you want a global cache
#set to file.path("FMBN", "FMBN_plus.rds") if you want a local cache in the project folder

FMBN_plus <- get_FMBN(use_local = p$use_local, force_download = p$force_download,
    do_cache = p$do_cache, cache_max_days = 180, local_FMBN_path = local_input_path
    )

# output data folder
if (!dir.exists(p$out_folder)) dir.create(p$out_folder)
```
`r FMBN_plus$version`.
I will focus on the `r p$taxon_level` level in this example and will go down to `r p$go_down_to`: technically, it would be better to include two or more genera which share the same niches, for comparison purposes.

```{r get-data, message = F, warning = F}

# collapse vector to a single label for use in text and captions
taxon_label <- str_c(p$taxa_to_search, collapse = ", ")

# check number of taxa requested
if (length(p$taxa_to_search) > 12) {
  warning("More than 12 taxa requested. Plots may be unreadable. ",
          "Consider reducing taxa_to_search to 12 or fewer.", call. = FALSE)
}

match_tax_level <- p$taxon_level %in% colnames(FMBN_plus$taxa)[3:8]
match_go_down   <- p$go_down_to  %in% colnames(FMBN_plus$taxa)[3:9]

if (!(match_tax_level & match_go_down)) cat(
  "\nWARNING: wrong taxonomic level(s)! Must be one of:\n",
  colnames(FMBN_plus$taxa)[3:9], "\n"
)

# match_go_down must be a taxonomic level below taxon_level
right_tax_order <- which(p$go_down_to   == colnames(FMBN_plus$taxa))[[1]] >
                   which(p$taxon_level  == colnames(FMBN_plus$taxa))[[1]]

# --- auto-detect domain from taxa_to_search --------------------------------
# look up each taxon in p$taxa_to_search across ALL taxa (both domains)
taxa <- FMBN_plus$taxa

detected_domains <- auto_detect_domain(taxon_to_search = p$taxa_to_search,
                                       taxon_level = p$taxon_level, 
                                       taxa_all = taxa)
# set domain automatically — overrides p$domain if present
if(detected_domains != p$domain) p$domain <- detected_domains

if (p$verbose) message("Domain auto-detected as: ", p$domain)

# --- taxonomic filtering ----------------------------------------------------

my_taxa <- taxa |>
  dplyr::filter(domain == p$domain)

unique_taxa <- unique(my_taxa[[p$taxon_level]])
unique_taxa <- unique_taxa[!is.na(unique_taxa)]

# %chin% is provided by data.table and is faster than %in% for character vectors
is_valid_taxon <- any(p$taxa_to_search %chin% unique_taxa)

my_taxa_in_FMBN <- my_taxa_in_FMBN <- unique_taxa[unique_taxa %chin% p$taxa_to_search]

is_valid_taxon <- all(my_taxa_in_FMBN %chin% unique_taxa)
if (!is_valid_taxon) stop(paste("\nWARNING: can't find your taxon in", p$taxon_level, "\n"))

if (length(p$taxa_to_search) == 1) {
  my_taxa <- dplyr::filter(my_taxa, .data[[p$taxon_level]] == p$taxa_to_search)
} else {
  my_taxa <- dplyr::filter(my_taxa, .data[[p$taxon_level]] %in% p$taxa_to_search)
}

# counts of taxa at taxon_level
taxa_count <- my_taxa |> dplyr::count(.data[[p$taxon_level]])

# optionally remove NA in the level immediately below taxon_level
tax_level_below <- case_when(
  p$taxon_level == "order"  ~ "family",
  p$taxon_level == "family" ~ "genus",
  p$taxon_level == "genus"  ~ "species"
)

my_taxa_all <- my_taxa

if (!p$keep_NA_lowest) my_taxa <- dplyr::filter(
  my_taxa, !is.na(.data[[tax_level_below]])
)

# --- edges ------------------------------------------------------------------
if (p$domain == "Bacteria") {
  edges <- FMBN_plus$edges_B
} else {
  edges <- FMBN_plus$edges_F
}

if (p$remove_chloroplast & p$domain == "Bacteria") {
  chloroplast_id <- FMBN_plus$taxa |>
    dplyr::filter(label == "Chloroplast" | class == "Chloroplast") |>
    pull(taxonId)
  edges <- dplyr::filter(edges, !(taxonId %in% chloroplast_id))
}

if (p$remove_mitochondria & p$domain == "Bacteria") {
  mitochondria_id <- FMBN_plus$taxa |>
    dplyr::filter(label == "Mitochondria" | family == "Mitochondria") |>
    pull(taxonId)
  edges <- edges |> dplyr::filter(!(taxonId %in% mitochondria_id))
}

if (p$remove_no_phylum) {
  no_phylum_id <- taxa |> dplyr::filter(is.na(phylum)) |> pull(taxonId)
  edges <- edges |> dplyr::filter(!(taxonId %in% no_phylum_id))
}

# re-calculate weight if any taxa were removed
if (any(p$remove_chloroplast, p$remove_mitochondria, p$remove_no_phylum)) {
  edges <- edges |>
    mutate(weight_old = weight) |>
    group_by(sampleId) |>
    mutate(weight = weight / sum(weight)) |>
    ungroup()
}

# note: weight is now a fraction; weight_old is the uncorrected value

n_edges <- nrow(edges)
# filtering join: keep only edges for taxa of interest
edges_filt <- semi_join(edges, my_taxa, by = "taxonId")

prop_edges_my_taxa <- nrow(edges_filt) / n_edges

# join taxonomy
edges_filt <- left_join(edges_filt,
                        select(my_taxa, taxonId:species),
                        by = "taxonId")

# --- handle NA at go_down_to level ------------------------------------------
tax_levels <- c("domain", "phylum", "class", "order", "family", "genus", "species")
target_level <- p$go_down_to
target_idx   <- match(target_level, tax_levels)

if (target_idx > 1) {
  parent_level <- tax_levels[target_idx - 1]
  edges_filt <- edges_filt |>
    mutate(!!sym(target_level) := case_when(
      !is.na(!!sym(target_level))  ~ !!sym(target_level),
      target_level == "species"    ~ str_c(!!sym(parent_level), " spp."),
      TRUE                         ~ !!sym(parent_level)
    ))
}

# --- pool edges by sample x go_down_to --------------------------------------
pooled_edges <- edges_filt |>
  group_by(sampleId, .data[[p$go_down_to]]) |>
  summarise(weight = sum(weight), .groups = "drop") |>
  arrange(sampleId, .data[[p$go_down_to]])

# --- samples ----------------------------------------------------------------
if (p$domain == "Bacteria") {
  filt_samples <- FMBN_plus$samples_B
} else {
  filt_samples <- FMBN_plus$samples_F
}

# remove blanks, mocks, and water samples
filt_samples <- filt_samples |>
  dplyr::filter(
    !str_detect(L1, "blank") &
    !str_detect(L1, "Mock")  &
    !str_detect(L1, "^Water")
  )

if (p$s_type %in% c("Environment", "Sample")) {
  filt_samples <- filt_samples |> dplyr::filter(s_type == p$s_type)
}

food_cat_all <- filt_samples |>
  distinct(.data[[p$food_grouping_variable]])

# inner join: also acts as a filtering join removing unwanted samples
pooled_edges_sample <- inner_join(
  pooled_edges,
  select(filt_samples, studyId, sampleId, s_type, n_reads2, n_issues,
         foodId, L1:target2, geo_loc_country),
  by = "sampleId"
)

food_cat_w_taxa <- pooled_edges_sample |> 
  distinct(.data[[p$food_grouping_variable]]) 

# --- studies ----------------------------------------------------------------
filt_studies <- FMBN_plus$studies

# join study info
pooled_edges_sample_study <- inner_join(
  pooled_edges_sample,
  select(filt_studies, studyId, read_length_bp, Seq_accn,
         food_group, overlapping, paired_end),
  by = "studyId"
)

food_cats_message <- setdiff(food_cat_all[[p$food_grouping_variable]], 
                             food_cat_w_taxa[[p$food_grouping_variable]])
if(length(food_cats_message)>0) {
  food_cats_message <- str_c("The food categories in which the taxa are NOT found are: ",
                           str_flatten_comma(food_cats_message),".", sep ="")
} else {
    food_cats_message <- ""
  }

# summary counts for in-line text
nstudies_wtarget <- pooled_edges_sample_study |> distinct(studyId) |> nrow()
nsamples_wtarget <- pooled_edges_sample_study |> distinct(sampleId) |> nrow()
nstudies_total   <- nrow(FMBN_plus$studies)
nsamples_total   <- if (p$domain == "Bacteria") {
  nrow(FMBN_plus$samples_B)
} else {
  nrow(FMBN_plus$samples_F)
}

```


`r nstudies_wtarget` studies include `r taxon_label` out of `r nstudies_total`. `r taxon_label` is found in `r nsamples_wtarget` samples out of `r nsamples_total`. The taxa were found in `r nrow(food_cat_all)` out of `r nrow(food_cat_w_taxa)`. `r food_cats_message`
  


## Where are my taxa?

I am now creating a table with the number of samples by food category which contain the taxon of interest. Note that, in order to calculate prevalence, I need a denominator. This really depends on the level of aggregation. Additionally, it would be better to perform some sort of sample equalization by group. Here, I will use a simplified approach and get the denominator I will use the following procedure:  

* select samples for the food group variable selected in parameters in which the target taxon/taxa is/are present; 

* get the total number of samples for each group 

* if the grouping variable is L1, get the L4 groups in which the target taxon/taxa are found; get the total number of samples for each group  

* further filtering by target and sequencing depth is optionally applied


```{r tbl-prevab_1, warning = F}

# remove any NA deriving from the left_join of samples
pooled_edges_sample_study <- pooled_edges_sample_study |>
  dplyr::filter(!is.na(.data[[p$food_grouping_variable]]))

# remove samples below minimum sequencing depth
# done here so all downstream outputs (prevalence, abundance, plots,
# specialisation) are consistent with each other
n_before_depth <- n_distinct(pooled_edges_sample_study$sampleId)
pooled_edges_sample_study <- pooled_edges_sample_study |>
  dplyr::filter(n_reads2 >= p$min_n_reads)
n_after_depth <- n_distinct(pooled_edges_sample_study$sampleId)

if (p$verbose) message(
  n_before_depth - n_after_depth, " samples removed (n_reads2 < ", p$min_n_reads, "); ",
  n_after_depth, " retained."
)

pooled_edges_sample_study_all <- pooled_edges_sample_study

if (p$run_further_filters) {
  pooled_edges_sample_study <- pooled_edges_sample_study |>
    dplyr::filter(target2 %in% p$gene_targets)
}

n_samples <- filt_samples |>
  dplyr::filter(sampleId %in% unique(pooled_edges_sample_study$sampleId)) |>
  summarise(n = n_distinct(sampleId), .by = all_of(p$food_grouping_variable))

# join original info and look at proportions
n_samples_all <- filt_samples |>
  dplyr::filter(s_type == p$s_type) |>
  summarise(n = n_distinct(sampleId), .by = all_of(p$food_grouping_variable)) |>
  dplyr::rename(n_all = n)

n_samples <- left_join(n_samples, n_samples_all, by = p$food_grouping_variable) |>
  mutate(prev = n / n_all) |>
  dplyr::arrange(desc(prev))

if (p$food_grouping_variable == "L1") {
  n_samples_all_L4 <- filt_samples |>
    dplyr::filter(s_type == "Sample") |>
    summarise(n = n_distinct(sampleId), .by = "L4") |>
    dplyr::rename(n_all = n)
}


if (p$save_data) {
  out_fn <- file.path(p$out_folder, str_c(p$out_fn_pref, "n_samples_table.txt"))
  write_tsv(n_samples, file = out_fn)
}

n_samples |>
  kable(caption = str_c("Number of samples in which ", taxon_label, " is found after filtering steps."),
        digits = 2) |>
  kable_classic()
```
Here is a table with studies and references.

```{r tbl-study-refs}
# get study info and references 
my_studies <- distinct(pooled_edges_sample_study, studyId) |>
  pull(studyId)
my_studies_df <- filt_studies |>
  dplyr::filter(studyId %in% my_studies) |>
  select(studyId, target, region, bioproject, food_group, DOI, ref_complete)
if (p$save_data) {
  write_tsv(my_studies_df, file.path(p$out_folder, str_c(p$out_fn_pref, "studies.txt")))
}
kable(my_studies_df, caption = str_c(
  "Details on FoodMicrobionet studies in which ",
  taxon_label, " are found."
)) |>
  kable_styling() |>
  scroll_box(width = "100%", height = "300px")
```

## Calculate prevalence and abundance.  

I will now calculate prevalence and abundance. More options (including more indices, see the filter_and_profile taxa function) should be added in the future.

```{r tbl-prevab-2}
# calculate prevalence and abundance 

prev_df <- pooled_edges_sample_study |>
  group_by(across(all_of(c(p$go_down_to, p$food_grouping_variable)))) |>
  summarise(abs_prev = n_distinct(sampleId),
            .groups = "drop_last") |>
  ungroup() |>
  left_join(n_samples, by = p$food_grouping_variable) |>
  mutate(prev = abs_prev/n_all)
# by default there are no 0 weight in FoodMicrobionet, they only become 0 if you pivot_wider the edges
# I could use an option for using log(abundance) and/or for choosing the base of the log (2 or 10) 

ab_df <- pooled_edges_sample_study |>
  mutate(log_ab = log(weight, base = p$log_base)) |>
  group_by(across(all_of(c(p$go_down_to, p$food_grouping_variable)))) |>
  summarise(min_ab = min(weight),
            max_ab = max(weight),
            mean_ab = mean(weight),
            median_ab = median(weight),
            mad_ab = mad(weight),
            median_logab = median(log_ab),
            .groups = "drop_last") |>
  ungroup()

prev_ab_df <- left_join(prev_df, ab_df, by = c(p$go_down_to, p$food_grouping_variable)) |>
  dplyr::arrange(p$go_down_to, p$food_grouping_variable)

if (p$save_data) {
  write_tsv(prev_ab_df, file.path(p$out_folder, str_c(p$out_fn_pref, "prev_ab_df.txt")))
}

kable(prev_ab_df, "html",
      caption = str_c(
        "Prevalence and abundance summary for ",
        taxon_label, "."
      ), digits = 3) |>
  kable_styling() |>
  scroll_box(width = "100%", height = "300px")

  
```

## Plotting the distributions.  

Here is a diagnostic plot on sequence length (shorter sequences can be optionally removed).

```{r fig-seq-length-plot, message = F, warning = F}
ggplot(pooled_edges_sample_study, mapping = aes(x=read_length_bp)) +
  facet_wrap(~target2) +
  geom_histogram()

# removing shorter sequences (only ifp$domain == "Bacteria)
if(p$domain == "Bacteria"){
  pooled_edges_sample_study <- pooled_edges_sample_study |>
    dplyr::filter(read_length_bp >= p$min_read_length)
  }

# how many studies left?
residual_studies <- unique(pooled_edges_sample_study$studyId)
# how many samples left?
residual_samples <- unique(pooled_edges_sample_study$sampleId)

# recalculate prev and ab
prev_df_sel <- pooled_edges_sample_study |>
  group_by(across(all_of(c(p$go_down_to, p$food_grouping_variable)))) |>
  summarise(abs_prev = n_distinct(sampleId),
            .groups = "drop_last") |>
  ungroup() |>
  left_join(n_samples, by = p$food_grouping_variable) |>
  mutate(prev = abs_prev/n_all)

# by default there are no 0 weight in FoodMicrobionet, they only become 0 if you pivot_wider the edges
# I could use an option for using log(abundance) and/or for choosing the base of the log (2 or 10) 

ab_df_sel <- pooled_edges_sample_study |>
  mutate(log_ab = log(weight, base = p$log_base)) |>
  group_by(across(all_of(c(p$go_down_to, p$food_grouping_variable)))) |>
  summarise(min_ab = min(weight),
            max_ab = max(weight),
            mean_ab = mean(weight),
            median_ab = median(weight),
            mad_ab = mad(weight),
            median_logab = median(log_ab),
            .groups = "drop_last") |>
  ungroup()


prev_ab_df_sel <- left_join(prev_df_sel, ab_df_sel, 
                            by = c(p$go_down_to, p$food_grouping_variable)) |>
  dplyr::arrange(p$go_down_to, p$food_grouping_variable)

if (p$save_data) {
  write_tsv(prev_ab_df_sel, file.path(p$out_folder, str_c(p$out_fn_pref, "prev_ab_df_sel.txt")))
}

```

After further filtering, there are `r length(residual_studies)` studies and `r length(residual_samples)` samples left.  


# Visualisations for prevalence and abundance.

I am using variations of box plots / violin plots and dotplots for abundance with colour clues for prevalence. Highlighting fermented and/or spoiled samples is also possible.

## Abbreviations for foods and taxa.  

```{r tbl-taxa-abbreviations}
# create abbreviations
# resolve_abbr, in lieu of str_trunc ensures that each abbreviation is unique
# need to use %>% as a pipe
df_to_plot <- pooled_edges_sample_study %>% 
  mutate(taxon_abbr = resolve_abbr(vals = .data[[p$go_down_to]], 
                                   width = p$abbr)$abbr,
         food_group_abbr = resolve_abbr(vals = .data[[p$food_grouping_variable]], 
                                     width = p$abbr)$abbr,
         log_ab = log(weight, base = p$log_base)) 

# I am transforming species into an ordered factor, to make plotting more 
# coherent with phylogenetic relatedness. 
# you may choose alternative sorting approaches here
species_levels <- sort(unique(pooled_edges_sample_study[[p$go_down_to]]))
taxa_abbr_levels <- sort(unique(df_to_plot$taxon_abbr))


df_to_plot <- df_to_plot |>
  mutate(taxon_abbr = factor(taxon_abbr, levels = taxa_abbr_levels),
         !!sym(p$go_down_to) := factor(.data[[p$go_down_to]], levels = species_levels))
# left join the prevalence data, by food group and taxon
df_to_plot <- df_to_plot |> 
  left_join(select(prev_ab_df_sel, all_of(c(p$go_down_to, p$food_grouping_variable)), rel_prev = prev))  
  
# the ordering of the factors could be adjusted
keys_taxa <- df_to_plot |> distinct(.data[[p$go_down_to]], taxon_abbr)
keys_taxa |> kable(caption = "Abbreviations for taxa names in plots.") |>
  kable_classic()
```


```{r tbl-food-abbreviations}
keys_foods <- df_to_plot |> distinct(.data[[p$food_grouping_variable]], food_group_abbr)
keys_foods |> kable(caption = "Abbreviations for food categories in plots") |>
  kable_classic()
```

## The distribution plot, finally.

The plot layout and style are controlled by several parameters in `p`:

* `plot_type` can be "boxplot" or "violin"  

* `violin_min_n` fall back to boxplot if any group has fewer obs

* `plot_facet_by`: `"food"` puts taxa on the x-axis and food categories on facets; `"taxon"` flips this.
* `plot_fill_var`: `"rel_prev"` maps relative prevalence (a group-level summary) to the box fill (continuous viridis); `"spoilage"` maps the fermentation/spoilage status (a sample-level property) to the colour of the jitter points; `"none"` leaves both plain.
* `plot_show_jitter`: if `TRUE`, individual samples are shown as grey points behind the boxes (outlier markers are automatically suppressed to avoid double-plotting).
* `plot_notch`: whether to notch the boxes.

```{r fig-boxplot, fig.height = 7, warning = FALSE, message = FALSE}

# override the colours defined in make_dist_plot if you want
override_make_dist_plot <- F # set this to T to change
if(override_make_dist_plot){
  fermspoil_scale_1 <- c("blue", "red", "cyan", "mediumorchid")
  # alternatives to consider:
  # High contrast, universally accessible
  fermspoil_scale_2 <- c("#E69F00", "#56B4E9", "#009E73", "#CC79A7")
  # Punchy and highly differentiable
  fermspoil_scale_3 <- c("#648FFF", "#785EF0", "#DC267F", "#FE6100") 
  # (Blue, Violet, Magenta, Orange)
  # (Orange, Sky Blue, Bluish Green, Reddish Purple)
  # Evenly spaced across the Viridis spectrum
  fermspoil_scale_4 <- c("#440154", "#31688E", "#35B779", "#FDE725") 
  # (Deep Purple, Muted Blue, Vibrant Green, Bright Yellow)
  fermspoil_scale <- fermspoil_scale_2# set this to change the palette
  names(fermspoil_scale) <- c("Unspoiled", "Spoiled", "Fermented", "Fermented+Spoiled")
}

weightlabel <- if_else(p$log_base == 2,
                       "log2(rel. abundance)",
                       "log10(rel. abundance)")

box_plot_out <- make_distplot(df_to_plot, p, weightlabel)
box_plot_out

if (p$save_graph) {
  out_fn <- file.path(p$out_folder, str_c(p$out_fn_pref, "prev_ab_boxplot", p$graph_type))
  ggsave(box_plot_out, filename = out_fn, dpi = p$graph_res, width = p$graph_width,
         height = p$graph_height)
}
```


### Dot plot: median abundance × prevalence.

```{r fig-dotplot, fig.height = 7, warning = FALSE, message = FALSE}

dot_plot_out <- make_dotplot(prev_ab_df_sel, p, weightlabel)
dot_plot_out

if (p$save_graph) {
  out_fn <- file.path(p$out_folder, str_c(p$out_fn_pref, "dotplot", p$graph_type))
  ggsave(dot_plot_out, filename = out_fn, dpi = p$graph_res, width = p$graph_width,
         height = p$graph_height)
}
```

# Specialisation.

This section characterises the degree of food-category specialisation of each taxon using
the d' index from the `bipartite` package. Note that the analyses in this
section apply additional filtering (see parameters `min_ab_specialisation` and
`min_prev_specialisation`) on top of the depth filter already applied to
`pooled_edges_sample_study`. Results here are therefore based on a subset of the data
and should be interpreted accordingly. The additional filtering removes taxon x food
category combinations supported by very few samples or very low abundance, which would
produce unreliable specialisation estimates.

## Diagnostic: sequencing depth by food category

This table summarises sequencing depth of the retained samples by food category. Food
categories with high `pct_below_threshold` had many shallow samples removed by the
depth filter and their prevalence and specialisation estimates should be interpreted
with caution.

```{r tbl-depth-diagnostics, warning = FALSE, message = FALSE}

depth_diag <- pooled_edges_sample_study |>
  distinct(sampleId, .data[[p$food_grouping_variable]], n_reads2) |>
  group_by(.data[[p$food_grouping_variable]]) |>
  summarise(
    n_samples           = n(),
    median_depth        = median(n_reads2, na.rm = TRUE),
    min_depth           = min(n_reads2,    na.rm = TRUE),
    max_depth           = max(n_reads2,    na.rm = TRUE),
    pct_below_threshold = mean(n_reads2 < p$min_n_reads, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(median_depth)

if (p$verbose) {
  n_shallow_cats <- sum(depth_diag$pct_below_threshold > 0.5)
  if (n_shallow_cats > 0)
    message(n_shallow_cats, " food category/ies have >50% of samples below the ",
            p$min_n_reads, " read threshold — interpret d' for these with caution.")
}

depth_diag |>
  kable(caption = str_c(
    "Sequencing depth by ", p$food_grouping_variable,
    " after depth filtering (min_n_reads = ", p$min_n_reads, "). ",
    "pct_below_threshold: proportion of samples originally below threshold ",
    "(shown for reference only — those samples are already excluded)."
  ), digits = 3) |>
  kable_classic() |>
  scroll_box(width = "100%", height = "300px")
```

## Build the interaction matrix for specialisation

```{r tbl-specialisation-filter, warning = FALSE, message = FALSE}

# step 1: compute depth-weighted or standard prevalence and median abundance
# per taxon x food category combination
# note: depth filtering is already applied to pooled_edges_sample_study
spec_summary <- pooled_edges_sample_study |>
  group_by(.data[[p$go_down_to]], .data[[p$food_grouping_variable]]) |>
  summarise(
    abs_prev      = if (p$depth_weighted_prev) {
      sum(pmin(1, n_reads2 / p$min_n_reads), na.rm = TRUE)
    } else {
      n_distinct(sampleId)
    },
    n_samples_grp = n_distinct(sampleId),
    median_ab     = median(weight,                  na.rm = TRUE),
    median_logab  = median(log(weight, p$log_base), na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(
    dplyr::select(n_samples, all_of(p$food_grouping_variable), n_all),
    by = p$food_grouping_variable
  ) |>
  mutate(rel_prev = abs_prev / n_all)

# step 2: apply abundance and prevalence thresholds
spec_summary_filt <- spec_summary |>
  dplyr::filter(
    median_ab     >= p$min_ab_specialisation,
    n_samples_grp >= p$min_prev_specialisation
  )

n_combos_before <- nrow(spec_summary)
n_combos_after  <- nrow(spec_summary_filt)

if (p$verbose) message(
  n_combos_before - n_combos_after, " taxon x food combinations dropped by ",
  "abundance/prevalence thresholds; ", n_combos_after, " retained."
)

# step 3: pivot to wide interaction matrix
# rows = taxa, columns = food categories
# weight is either rel_prev or median_ab — log-abundance is not used here
# because d' marginals require weights with a proportional interpretation
weight_var <- p$bipartite_weight_var  # "rel_prev" or "median_ab"

interact_matrix <- spec_summary_filt |>
  dplyr::select(all_of(c(p$go_down_to, p$food_grouping_variable, weight_var))) |>
  pivot_wider(
    names_from  = all_of(p$food_grouping_variable),
    values_from = all_of(weight_var),
    values_fill = 0
  ) |>
  column_to_rownames(p$go_down_to)

if (p$verbose) message(
  "Interaction matrix: ", nrow(interact_matrix), " taxa x ",
  ncol(interact_matrix), " food categories. ",
  "Weight variable: ", weight_var, "."
)

interact_matrix |>
  round(3) |>
  kable(caption = str_c(
    "Interaction matrix used for specialisation analysis. ",
    "Weight variable: ", weight_var, ". ",
    "Zeros indicate taxon x food combinations excluded by filtering."
  )) |>
  kable_classic() |>
  scroll_box(width = "100%", height = "300px")
```

## Specialisation index d'

d' measures how much each taxon concentrates its interactions on a subset of food
categories relative to what would be expected from the marginal distributions of the
network. d' = 0 indicates a perfect generalist (taxon uses all food categories in
proportion to their overall availability); d' = 1 indicates a perfect specialist.
Values should be interpreted cautiously for taxa present in few food categories or
for food categories with high depth heterogeneity (see diagnostic table above). 
Note however that FoodMicrobionet is not a random sampling of foods: some foods 
may be over represented and other under-represented. Also, the foods that are 
included depend from the ensemble of taxa you have selected: The more taxa you use 
the likelier it is that you include a wide variety of foods and that the specialisation.  
indices are meaningful.  

```{r tbl-specialisation-dprime, warning = FALSE, message = FALSE}

spec_result <- bipartite::specieslevel(
  t(interact_matrix),
  level = "higher",  # taxa are rows = higher trophic level in bipartite terminology
  index = "d"
)

dprime_df <- spec_result |>
  as_tibble(rownames = p$go_down_to) |>
  rename(d_prime = d) |>
  arrange(desc(d_prime))

if (p$verbose) message(
  "d' computed for ", nrow(dprime_df), " taxa across ",
  ncol(interact_matrix), " food categories."
)

# identify taxa dropped from d' due to insufficient data
taxa_in_data   <- unique(pooled_edges_sample_study[[p$go_down_to]])
taxa_in_dprime <- dprime_df[[p$go_down_to]]
taxa_dropped_dprime <- setdiff(taxa_in_data, taxa_in_dprime)

d_prime_message <- if(length(taxa_dropped_dprime) > 0) str_c("The following taxa were excluded because they did not meet the minimum abundance (", p$min_ab_specialisation, ") or prevalence (", p$min_prev_specialisation, " samples) thresholds in any food category: ", str_flatten_comma(taxa_dropped_dprime), ".") else "All taxa met the thresholds for d' calculation."

dprime_df |>
  kable(caption = str_c(
    "Specialisation index d' for ", taxon_label, ". ",
    "d' = 0: perfect generalist; d' = 1: perfect specialist. ",
    "Weight variable used: ", weight_var, "."
  ), digits = 3) |>
  kable_classic()
```

d' was computed for `r nrow(dprime_df)` out of `r length(unique(pooled_edges_sample_study[[p$go_down_to]]))` taxa.
`r d_prime_message`

# Conclusions.  

These are mostly up to you, the user of this script. Consider carefully what you want to use: one taxon? Two or more? At which taxonomic resolution?

# Acknowledements.  

Claude AI (Sonnet 4.6) was used to 

1. suggest the initial version of the dot plot and bipartite analysis which were then revised by me and polished by Claude in a second pass  

2. more generally, to revise the code


# Copyright. 

This is version 1.2.5 (2026-06-06).

Assume that this is overall under MIT licence.

Copyright 2026 Eugenio Parente.  
Permission is hereby granted, free of charge, to any person obtaining a copy of this software and associated documentation files (the "Software"), to deal in the Software without restriction, including without limitation the rights to use, copy, modify, merge, publish, distribute, sublicense, and/or sell copies of the Software, and to permit persons to whom the Software is furnished to do so, subject to the following conditions:
THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.

# References.  
